Beyond Static Retrieval: Adaptive Method for Improving RAG Performance

### 2025-03-08 Paper/latex/paper_RAG.tex
last update

\abstract{Retrieval-Augmented Generation (RAG) improves the quality of language model responses by incorporating external knowledge. However, using a fixed number of retrieved documents (K) fails to adapt to varying query complexity, often leading to suboptimal retrieval either including irrelevant documents or missing crucial ones. To address this issue, we propose a hybrid approach that dynamically adjusts K based on query characteristics while maintaining computational efficiency. Our method improves retrieval performance by maximizing relevant information and minimizing noise. Experimental results show that it outperforms fixed-K retrieval, offering a more effective solution for optimizing RAG systems. Future work will explore learning-based techniques to further refine dynamic K selection. Future work will explore learning-based techniques to further refine dynamic K selection.}

@@

Static Top-K selection in Retrieval :In traditional Retrieval-Augmented Generation (RAG) systems, static K-selection is a widely used method where a fixed number of top-k documents are retrieved based on their relevance scores. Sparse retrieval\cite{bai2020spartermlearningtermbasedsparse} methodsw here documents and queries are represented as high-dimensional, sparse vectors In these method , the "keys" represent documents to be retrieved, and the "values" are he importance or weight of those terms in the document such as query likelihood models\cite{Lafferty2001} ,TF-IDF\cite{Robertson1997} and BM25\cite{Zaragoza2009} often rely on this static approach to identify and rank documents. For instance, if k  =5, the system retrieves the top 5 documents deemed most relevant to a query, regardless of the query's complexity or the variability in relevance distribution among documents.
A major drawback of this approach is its rigidity. Since the number of retrieved documents (k) is fixed, it fails to adapt to the varying complexity of queries. For simple queries, retrieving many documents may introduce noise, while for complex queries, retrieving few documents may result in incomplete or insufficient information. This one-size-fits-all strategy can lead to suboptimal performance across diverse datasets or tasks.

Dynamic top-k Selection in Retrieval :
Dynamic k-selection addresses the limitations of static k by varying the number of retrieved documents based on the specific query or context. These approaches typically involve heuristics, machine learning models, or adaptive algorithms to determine the optimal K for each query such as Dynamic Query Cutoff (DQC)\cite{Culpepper2016} This approach uses machine learning models trained on query-document features to predict the optimal k for retrieval , A reinforcement learning-based approach\cite{Zhou2021} where the agent learns to adjust k by maximizing downstream task performance (e.g accurate response generation in RAG) Another example is the Dynamic Selection of k Nearest Neighbors \cite{DynamicKNN2012} where the selection process often depends on the distribution of neighbors, their distances to the query point, or their relevance scores.
However, these methods come with limitations.They often introduce computational overhead due to the need for dynamic evaluation, which can increase latency in large-scale systems. Additionally, their performance heavily depends on the quality of the underlying models or distance metrics, making them sensitive to noisy data or poorly tuned hyperparameters. As a result, achieving generalization across diverse datasets or domains remains a significant challenge.

Hybrid Approaches For Top-K selection in Retrieval :
combine elements of static and dynamic K-selection to achieve a balance between simplicity and adaptability.These systems might use a static 
k as a baseline but dynamically adjust such as STAYKATE (Static-Dynamic Hybrid Selection)\cite{Zhu2024} Combines static (active learning-inspired) and dynamic (KATE-based) example selection strategies to create STAYKATE. Another example is DR-RAG (Dynamic Relevance for Retrieval-Augmented Generation)\cite{hei2024drragapplyingdynamicdocument} retrieval framework designed to improve the relevance between queries and documents. In the first stage, a similarity matching method retrieves an initial set of documents. In the second stage, a deeper analysis is performed to identify dynamic-relevant documents, supported by a classifier that filters out irrelevant ones . however hybrid methods may face challenges in determining when to apply static or dynamic k-selection effectively, leading to potential trade-offs between retrieval accuracy and system efficiency. These limitations highlight the ongoing need for robust mechanisms that dynamically and efficiently adapt k to the unique characteristics of each query and dataset.

>>>>

\abstract{Retrieval-Augmented Generation (RAG) improves the quality of language model responses by incorporating external knowledge. However, using a fixed number of retrieved documents (K) fails to adapt to varying query complexity, often leading to suboptimal retrieval either including irrelevant documents or missing crucial ones. To address this issue, we propose a hybrid approach that dynamically adjusts K based on query characteristics while maintaining computational efficiency. Our method improves retrieval performance by maximizing relevant information and minimizing noise. Experimental results show that it outperforms fixed-K retrieval, offering a more effective solution for optimizing RAG systems. Future work will explore learning-based techniques to further refine dynamic K selection.}

@@

Static Top-K selection in Retrieval :In traditional Retrieval-Augmented Generation (RAG) systems, static K-selection is a widely used method where a fixed number of top-k documents are retrieved based on their relevance scores. Sparse retrieval\cite{bai2020spartermlearningtermbasedsparse} methods where documents and queries are represented as high-dimensional, sparse vectors In these method , the "keys" represent documents to be retrieved, and the "values" are the importance or weight of those terms in the document such as query likelihood models\cite{Lafferty2001} ,TF-IDF\cite{Robertson1997} and BM25\cite{Zaragoza2009} often rely on this static approach to identify and rank documents. For instance, if k=5, the system retrieves the top 5 documents deemed most relevant to a query, regardless of the query's complexity or the variability in relevance distribution among documents.
A major drawback of this approach is its rigidity. Since the number of retrieved documents (k) is fixed, it fails to adapt to the varying complexity of queries. For simple queries, retrieving many documents may introduce noise, while for complex queries, retrieving few documents may result in incomplete or insufficient information. This one-size-fits-all strategy can lead to suboptimal performance across diverse datasets or tasks.

Dynamic top-k Selection in Retrieval :addresses the limitations of static k by varying the number of retrieved documents based on the specific query or context. These approaches typically involve heuristics, machine learning models, or adaptive algorithms to determine the optimal K for each query such as Dynamic Query Cutoff (DQC)\cite{Culpepper2016} This approach uses machine learning models trained on query-document features to predict the optimal k for retrieval , A reinforcement learning-based approach\cite{Zhou2021} where the agent learns to adjust k by maximizing downstream task performance (e.g accurate response generation in RAG) Another example is the Dynamic Selection of k Nearest Neighbors \cite{DynamicKNN2012} where the selection process often depends on the distribution of neighbors, their distances to the query point, or their relevance scores.
However, these methods come with limitations.They often introduce computational overhead due to the need for dynamic evaluation, which can increase latency in large-scale systems. Additionally, their performance heavily depends on the quality of the underlying models or distance metrics, making them sensitive to noisy data. As a result, achieving generalization across diverse datasets or domains remains a significant challenge.

Hybrid Approaches For Top-K selection in Retrieval :combine elements of static and dynamic K-selection to achieve a balance between simplicity and adaptability.These systems might use a static 
k as a baseline but dynamically adjust such as STAYKATE (Static-Dynamic Hybrid Selection)\cite{Zhu2024} Combines static (active learning-inspired) and dynamic (KATE-based) example selection strategies to create STAYKATE. Another example is DR-RAG (Dynamic Relevance for Retrieval-Augmented Generation)\cite{hei2024drragapplyingdynamicdocument} retrieval framework designed to improve the relevance between queries and documents. In the first stage, a similarity matching method retrieves an initial set of documents. In the second stage, a deeper analysis is performed to identify dynamic-relevant documents, supported by a classifier that filters out irrelevant ones . however hybrid methods may face challenges in determining when to apply static or dynamic k-selection effectively, leading to potential trade-offs between retrieval accuracy and system efficiency. These limitations highlight the ongoing need for robust mechanisms that dynamically and efficiently adapt k to the unique characteristics of each query and dataset.
